Application of LSTM Neural Networks and Complex Network Analysis for Earthquake Forecasting in Colombia

### 2025-02-27 article.tex
fix: revamp images
fix: improve introduction add reference to network feature concepts

As stated by León et al. \cite{leon_modeling_2018}, representing seismological events as complex networks allows for the study of seismicity in a region and helps to unravel the relationships among these events. Earthquake networks are based on the temporal sequence of events and their specific locations. Different network features, such as path length, degree distribution, and clustering, help define the relationships between events, hubs, and small-world network properties. Thus, integrating the analysis of complex networks to structure the relationships between seismic events seems to be helpful in improving forecast results.

In this study, we processed earthquake data from Colombia and trained LSTM neural network models. The models achieved 60-70\% accuracy for 2-quantile classification and similar accuracy for 3-quantile classification when using complex network features. These findings suggest that complex networks can enhance earthquake forecasting.

@@

\begin{figure}[H]
    \includegraphics[width=0.49\textwidth]{img/mag01.png}
    \includegraphics[width=0.49\textwidth]{img/depth01.png}
    \caption{Earthquake events visualization by Magnitude (left) and Depth (right)}\label{fig1}
\end{figure}
\unskip

@@

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{img/diagram-20250202.png}
        \caption{Quick overview of the creation of an earthquake network. A geographical region is divided into nodes of equal size in kilometers, then each event is clasified in each node by its latitude and longitude, we later connect each node by the time sequence of the events. We can observe nodes 7, 5, 8, 3 are connected in that order, first (7,5), then (5,8), and finally (8,3).
        } \label{fig2}
    \end{center}
\end{figure}
\unskip

@@

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.6\textwidth]{img/dataset-diagram.png}
        \caption{Dataset for the LSTM neural network.\label{fig3}}
    \end{center}
\end{figure}

>>>>

In his work, León et al. \cite{leon_modeling_2018} showed that representing seismological events as complex network helps understand of seismicity in a region and unravel the relationships among these events. Earthquake networks are based on the temporal sequence of events and their specific locations. Different network features, such as path length, degree distribution, and clustering, help define the relationships between events, hubs, and small-world network properties. For example, \textbf{Path length} measures the average shortest path between nodes in the network, indicating how quickly information (or seismic activity) can spread through the network \cite{newman_2010}. \textbf{Degree distribution} describes the probability distribution of the degrees over the entire network, helping to identify nodes with a high number of connections, which may act as critical points in the network \cite{barabsi_2009}. \textbf{Clustering coefficient} measures the degree to which nodes in a graph tend to cluster together, providing insights into the local interconnectedness of the network \cite{watts_1998}.

In this work, we wanted to answer whether incorporating network features can enhance the performance of LSTM models for earthquake forecasting. LSTM models have been widely used to model temporal data and earthquake forecasting \cite{lai_modeling_2018, cai_anomaly_2019,banna_attention-based_2021,wang_earthquake_2020, nazaria_predicting_2021}. By processing earthquake data from Colombia and training LSTM neural network models, we found that models achieved 60-70\% accuracy for 2-quantile classification and similar accuracy for 3-quantile classification when network features were included. These results indicate that complex network features can indeed improve earthquake forecasting accuracy.

@@

\begin{figure}[H]
    \includegraphics[width=0.49\textwidth]{img/mag.png}
    \includegraphics[width=0.49\textwidth]{img/depth.png}
    \caption{Earthquake events visualization by Magnitude (left) and Depth (right)}\label{fig1}
\end{figure}
\unskip

@@

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{img/eqnetwork.eps}
        \caption{Quick overview of the creation of an earthquake network. A geographical region is divided into nodes of equal size in kilometers, then each event is clasified in each node by its latitude and longitude, we later connect each node by the time sequence of the events. We can observe nodes 7, 5, 8, 3 are connected in that order, first (7,5), then (5,8), and finally (8,3).
        } \label{fig2}
    \end{center}
\end{figure}
\unskip

@@

\begin{figure}[H]
    \begin{center}
        \includegraphics[width=0.6\textwidth]{img/dataset.eps}
        \caption{Dataset for the LSTM neural network.\label{fig3}}
    \end{center}
\end{figure}
